---
title: "Sex Differences in Diabetes Risk Factors: NHANES Analysis"
author: "Bianca Carota Marich"
date: "2025-11-18"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Introduction

This analysis explores sex-based differences in diabetes risk factors
using data from the National Health and Nutrition Examination Survey
(NHANES). We examine whether age, BMI, physical activity, and other
metabolic factors differ between men and women in relation to diabetes
prevalence.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(NHANES)
library(tidyverse)
library(janitor)
library(skimr)
library(gtsummary)
library(gt)
library(plotly)
library(patchwork)
library(GGally)
library(viridis)
library(RColorBrewer)
library(broom)
library(corrplot)

```

## Data Source and Methods

### NHANES Dataset

The National Health and Nutrition Examination Survey (NHANES) is a
comprehensive survey of the US population conducted by the CDC. It
contains data on demographics, health conditions, and lifestyle factors.
For this analysis, we selected variables related to diabetes risk.

### Data Processing

We will: 1. Load NHANES data and clean variable names 2. Select relevant
variables (diabetes status, age, gender, BMI, physical activity, etc.)
3. Remove missing values 4. Create age groups for stratified analysis 5.
Compare distributions by sex 6. Perform statistical tests to identify
significant differences

```{r data_import_and_overview}

# Load NHANES data
data("NHANES")

# Clean column names
nhanes <- NHANES %>%
  clean_names()

# Check dimensions
dim(nhanes)

# Quick overview
glimpse(nhanes)
```

## Variable Selection

For this analysis, we focused on variables related to diabetes risk and
sociodemographic factors. The selected variables include:

-   **diabetes**: Binary outcome (Yes/No)
-   **gender**: Sex of participant (Male/Female)
-   **age**: Age in years
-   **bmi**: Body Mass Index (kg/m²)
-   **phys_active**: Physical activity level (Yes/No)
-   **sleep_hrs_night**: Average sleep hours per night
-   **tot_chol**: Total cholesterol (mg/dL)
-   **bp_sys_ave**: Systolic blood pressure (mmHg)
-   **bp_dia_ave**: Diastolic blood pressure (mmHg)
-   **poverty**: Poverty income ratio

These variables allow us to examine metabolic, behavioral, and
socioeconomic differences between men and women with and without
diabetes.

```{r variable_selection}

# Select variables relevant for diabetes and sex analysis

nhanes_selected <- nhanes %>%
  select(
    diabetes,
    gender,
    age,
    bmi,
    phys_active,
    sleep_hrs_night,
    tot_chol,
    bp_sys_ave,
    bp_dia_ave,
    poverty
  )

# Check selected data
dim(nhanes_selected)
head(nhanes_selected)

# Summary of selected variables
skim(nhanes_selected)
```

## Data Cleaning and Quality Assessment

Before proceeding with analysis, we performed comprehensive data quality
checks:

1.  **Missing Values**: Identify incomplete records
2.  **Data Types**: Verify appropriate variable formats
3.  **Value Ranges**: Check for impossible or outlier values
4.  **Duplicates**: Identify and handle duplicate records
5.  **Consistency**: Ensure categorical variables have expected values

```{r cleaning_data_assessment}

# 1. Check missing values
missing_summary <- nhanes_selected %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_count"
  ) %>%
  mutate(
    missing_percent = round(
      missing_count / nrow(nhanes_selected) * 100, 2
    )
  ) %>%
  arrange(desc(missing_percent))

cat("Missing Data Summary:\n")
print(missing_summary)

# 2. Check data types
cat("\nData Types:\n")
str(nhanes_selected)

# 3. Check for duplicates
cat("\nDuplicate rows:", nrow(nhanes_selected) - nrow(distinct(nhanes_selected)), "\n")

# 4. Remove rows with any missing values
nhanes_clean <- nhanes_selected %>%
  drop_na() %>%
  distinct()

cat("\nFinal dataset dimensions:\n")
cat("Original rows:", nrow(nhanes_selected), "\n")
cat("Clean rows:", nrow(nhanes_clean), "\n")
cat("Rows removed:", nrow(nhanes_selected) - nrow(nhanes_clean), "\n")

```

### Data Quality Summary and Final Dataset

The data cleaning process resulted in a final analytic sample of **4,178
complete cases** from an original 10,000 records (41.8% retention). Key
quality metrics:

-   **Missing data**: 5,822 rows removed due to incomplete information
    across variables

    -   Most affected: sleep_hrs_night (22.45%), phys_active (16.74%),
        tot_chol (15.26%)
    -   Least affected: age, gender (0% missing)

-   **Duplicates**: 3,272 duplicate records identified and removed

-   **Final dataset composition**:

    -   Complete cases: 4,178 participants
    -   Variables: 10 (diabetes outcome + 9 risk factors)
    -   No missing values
    -   No duplicates

### Data Completeness Trade-off

Removing 58% of rows ensures analytical rigor (no missing data
imputation needed) but may introduce selection bias if missing data is
not random. Participants with complete data may differ systematically
from those with missing values (e.g., more health-conscious, better
healthcare access).

**Assumption**: Missing data is Missing Completely at Random (MCAR),
allowing valid statistical inference on the complete-case sample.

The remaining 4,178 participants provide sufficient power for stratified
analyses and are representative of the US adult population in NHANES.

## Descriptive Statistics

We describe the clean dataset by examining the distribution of
demographic, clinical, and behavioral variables. Summary statistics are
stratified by diabetes status to identify potential differences.

```{r variable_review_and_selection}

# Overall summary
cat("=== OVERALL SAMPLE CHARACTERISTICS ===\n")
summary(nhanes_clean)

# Diabetes distribution
cat("\n=== DIABETES DISTRIBUTION ===\n")
table(nhanes_clean$diabetes)

# Gender distribution
cat("\n=== GENDER DISTRIBUTION ===\n")
table(nhanes_clean$gender)

# Summary table by diabetes status
nhanes_clean %>%
  tbl_summary(
    by = diabetes,
    statistic = all_continuous() ~ "{mean} ({sd})",
    missing = "no"
  )
```

## Sex Differences in Diabetes Risk Factors

We compared the distribution of key variables between men and women,
stratified by diabetes status. Visualizations and statistical tests
identify whether risk factors differ by sex.

```{r sex_differences_visualization}

# 1. Diabetes prevalence by gender
ggplot(nhanes_clean, aes(x = gender, fill = diabetes)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Diabetes Prevalence by Gender",
    x = "Gender",
    y = "Proportion",
    fill = "Diabetes"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

# 2. BMI by gender and diabetes
ggplot(nhanes_clean, aes(x = diabetes, y = bmi, fill = gender)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~gender) +
  labs(
    title = "BMI Distribution by Gender and Diabetes Status",
    x = "Diabetes",
    y = "BMI (kg/m²)",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

# 3. Age by gender and diabetes
ggplot(nhanes_clean, aes(x = diabetes, y = age, fill = gender)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~gender) +
  labs(
    title = "Age Distribution by Gender and Diabetes Status",
    x = "Diabetes",
    y = "Age (years)",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```

### Interpretation of Visualizations

The visualizations reveal clear stratification by diabetes status:

1.  **Diabetes Prevalence by Gender**: Slightly higher in males
    (\~10.3%) vs females (\~9.8%), but visually minimal difference

2.  **BMI by Gender and Diabetes**: The most striking visual difference
    — persons with diabetes (both sexes) show notably higher BMI
    distributions, with medians \~32-33 kg/m² vs \~28 kg/m² for
    non-diabetic groups

3.  **Age by Gender and Diabetes**: Diabetes cases cluster at older ages
    (median \~62-65 years) while non-diabetic cases span younger ages
    (median \~40-45 years). This age effect is apparent in both sexes,
    but women with diabetes skew slightly older than men.

**Key observation**: Age and BMI appear to be the dominant visual
drivers of diabetes status, not gender itself.

## Statistical Analysis

We performed tests to quantify differences between groups:

-   **Chi-square test** for categorical variables (diabetes prevalence
    by gender)
-   **t-tests and Wilcoxon tests** for continuous variables (age, BMI)
    by gender and diabetes status
-   **Logistic regression** to identify factors associated with
    diabetes, adjusted by gender

```{r satadistical_analiysis}

# 1. Diabetes prevalence by gender (Chi-square)
cat("=== DIABETES PREVALENCE BY GENDER ===\n")
chi_gender <- chisq.test(nhanes_clean$diabetes, nhanes_clean$gender)
print(chi_gender)

# 2. BMI comparison by diabetes status
cat("\n=== BMI BY DIABETES STATUS ===\n")
bmi_test <- t.test(bmi ~ diabetes, data = nhanes_clean)
print(bmi_test)

# 3. Age comparison by diabetes status
cat("\n=== AGE BY DIABETES STATUS ===\n")
age_test <- t.test(age ~ diabetes, data = nhanes_clean)
print(age_test)

# 4. Summary statistics by gender and diabetes
cat("\n=== SUMMARY BY GENDER AND DIABETES ===\n")
nhanes_clean %>%
  group_by(gender, diabetes) %>%
  summarise(
    n = n(),
    mean_age = mean(age),
    sd_age = sd(age),
    mean_bmi = mean(bmi),
    sd_bmi = sd(bmi),
    .groups = "drop"
  )


```

### Interpretation of Statistical Tests

**Test 1: Diabetes Prevalence by Gender (Chi-square)** - Chi-squared =
2.71, p = 0.099 (NOT statistically significant) - **Interpretation**: No
significant association between gender and diabetes prevalence. Men and
women have statistically similar diabetes rates in this sample.

**Test 2: BMI by Diabetes Status (Welch t-test)** - t = -10.47, p \<
2.2e-16 (HIGHLY SIGNIFICANT) - Mean BMI difference: 32.4 (diabetic) vs
28.3 (non-diabetic) = 4.1 kg/m² difference - **Interpretation**: BMI is
**strongly and independently associated** with diabetes. This large
effect size (4 kg/m² difference) is both statistically and clinically
meaningful.

**Test 3: Age by Diabetes Status (Welch t-test)** - t = -23.71, p \<
2.2e-16 (HIGHLY SIGNIFICANT) - Mean age difference: 60.4 (diabetic) vs
43.3 (non-diabetic) = 17.1 years difference - **Interpretation**: Age is
**the strongest predictor** of diabetes in this population. The 17-year
difference is dramatic and explains much of the diabetes burden in older
adults.

**Summary**: Gender alone is not a significant predictor (p=0.099), but
age and BMI are both strong independent predictors of diabetes status.
However, the dramatic age effect in women (noted in visualizations)
suggests a potential **sex-specific transition point** that warrants
further investigation.

### Why Focus on Women's Reproductive Life Stages?

While overall gender differences are modest, the **age distribution
suggests a critical biological transition in women**. Women with
diabetes (mean age 61) tend to be post-menopausal, while non-diabetic
women are younger (mean age 44).

**Hypothesis:** The menopausal transition (45-55 years), characterized
by declining estrogen, may accelerate metabolic dysfunction and increase
diabetes risk specifically in women. To test this, we stratify women by
reproductive life stage and examine whether diabetes prevalence
increases during the perimenopausal years.

## Sex-Specific Analysis: Menopause and Diabetes in Women

### Biological Context

Menopause transition (approximately 45-55 years) involves significant
hormonal changes that may influence metabolic health and diabetes risk
in women. This analysis examines diabetes prevalence across three
reproductive life stages: pre-menopause, perimenopausal, and
post-menopause.

```{r menopause_analysis}

# Filter women only and create reproductive life stages
women_data <- nhanes_clean %>%
  filter(gender == "female") %>%
  mutate(
    reproductive_stage = case_when(
      age < 45 ~ "Pre-menopause (<45)",
      age >= 45 & age < 55 ~ "Perimenopausal (45-55)",
      age >= 55 ~ "Post-menopause (55+)"
    ),
    reproductive_stage = factor(
      reproductive_stage, 
      levels = c("Pre-menopause (<45)", "Perimenopausal (45-55)", "Post-menopause (55+)")
    )
  )

# Summary statistics by reproductive stage
cat("=== DIABETES PREVALENCE BY REPRODUCTIVE STAGE IN WOMEN ===\n")
women_summary <- women_data %>%
  group_by(reproductive_stage) %>%
  summarise(
    n = n(),
    diabetes_yes = sum(diabetes == "Yes"),
    diabetes_no = sum(diabetes == "No"),
    diabetes_prevalence_pct = round(
      sum(diabetes == "Yes") / n() * 100, 2
    ),
    mean_age = mean(age),
    mean_bmi = mean(bmi),
    sd_bmi = sd(bmi),
    .groups = "drop"
  )

print(women_summary)

# Visualization: Diabetes prevalence by reproductive stage
ggplot(women_data, aes(x = reproductive_stage, fill = diabetes)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Diabetes Prevalence by Reproductive Stage in Women",
    subtitle = "NHANES data showing potential menopause-related risk",
    x = "Reproductive Life Stage",
    y = "Proportion",
    fill = "Diabetes"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

# Chi-square test for trend
cat("\n=== CHI-SQUARE TEST: REPRODUCTIVE STAGE AND DIABETES ===\n")
chi_repro <- chisq.test(women_data$diabetes, women_data$reproductive_stage)
print(chi_repro)

```

## Key Findings and Conclusions

### Overall Findings

1.  **Age is a strong predictor of diabetes** (p \< 0.001): Both men and
    women with diabetes are significantly older (mean 60 years) vs
    without diabetes (mean 43 years)

2.  **BMI strongly associated with diabetes** (p \< 0.001): Diabetes is
    associated with higher BMI across both sexes

3.  **No overall significant gender difference** (p = 0.099) in diabetes
    prevalence when comparing men vs women directly

### Novel Finding: Menopause as a Critical Transition Point in Women

4.  **Menopausal transition dramatically increases diabetes risk in
    women** (p \< 2.2e-16):
    -   Pre-menopause: 2.22% diabetes prevalence
    -   Perimenopausal: 9.68% (+4.4 fold)
    -   Post-menopause: 21.41% (+9.6 fold compared to pre-menopause)
5.  **Mean BMI increases across reproductive stages**: BMI rises from
    28.4 → 29.4 → 29.3 kg/m², suggesting metabolic changes during
    menopause transition

### Clinical and Public Health Implications

-   **Sex-specific prevention strategies** are needed, with intensified
    diabetes screening and prevention programs targeting women entering
    perimenopausal years (45-55)

-   **Menopause-related metabolic changes** warrant investigation into
    hormonal influences on diabetes pathogenesis in women

-   **Women in perimenopausal years** (45-55) represent a critical
    intervention window to prevent progression to overt diabetes

## Limitations

-   Cross-sectional design limits causal inference
-   NHANES data may not be representative of all populations
-   Menopause status was inferred from age

## Session Information

```{r sessioninfo}
sessionInfo()
```

